---
title: "TCX Analyzer"
author: ""
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The package supplies a set of functionality to analyze workouts TCX files. Such files can normally be exported from websites such as Garmin Connect or in some cases directly from the device recording it.

## Requirements

* A TCX containing at least time information. Although several TCX have been tested, the developer cannot guarantee 100% compatibility with all TCX files
* For TSS calculations a FTP/LHT is requested

# Quick reference

The following session provides a brief overview of the main package functions. To view functions details and a complete list of arguments, use the R command *??functionName*
```{r example0, echo=FALSE, warning=FALSE, message=FALSE}
library(tcxAnalyser)
gpx <- intervalActivity
```



## Load data
```{r example1, echo=TRUE, warning=FALSE, message=FALSE, eval=FALSE}
#Do not execute
gpx <- dataLoader('path_to_file.tcx')
```

## View a summary table
```{r example2, echo=TRUE, warning=FALSE, message=FALSE}
compareSplits(gpx)
```

## View a summary table and compute TSS
```{r example3, echo=TRUE, warning=FALSE, message=FALSE}
compareSplits(gpx, ftp = 4.1, ftpType = "pace")
```

## View a map
```{r example4, echo=TRUE, warning=FALSE, message=FALSE}
generateMap(gpx)
```

## Plot the activity metrics
```{r example5, echo=TRUE, warning=FALSE, message=FALSE}
doPlots(gpx,xVariable = "Time", showMe = colnames(gpx), doFacet = TRUE)
```

## Smooth data
```{r example5a, echo = TRUE, warning = FALSE, message=FALSE}
smoothedData <- gpxSmooth(gpx, variable = "Pace", k = 39)
#If variable is not given then all variables in gpxOriginal will be smoothed, except
#Time, Distance, LatitudeDegrees, LongitudeDegrees
smoothedData <- gpxSmooth(gpx, k = 39)
```
## Generate splits

### Manually
```{r example6, echo=TRUE, warning=FALSE, message=FALSE}
sp <- createSplits(gpx, 1500, "everyKm")
```

### Assisted
```{r example7, echo=TRUE, warning=FALSE, message=FALSE}
proposedSplits <- c(5, 10)
addSeparator(gpx, splitHere = proposedSplits, xVariable = "Time", yVariable = "Distance")

proposedSplits <- c(9, 13, 15, 19, 21.2, 25.1, 27.3, 31.4, 33.4, 37.5, 39.5)
addSeparator(gpx, splitHere = proposedSplits, xVariable = "Time", yVariable = "Distance")

sp <- createSplits(gpx, proposedSplits, "thisMin")
```

### Automatically
```{r example8, echo=TRUE, warning=FALSE, message=FALSE}
sp <- autoSplits(gpx)
```


## View a summary table and compute TSS
```{r example9, echo=TRUE, warning=FALSE, message=FALSE}
compareSplits(sp, ftp = 4.1, ftpType = "pace")

```

## View a map
```{r example10, echo=TRUE, warning=FALSE, message=FALSE}
generateMap(sp)
```

## Plot the activity metrics
```{r example11, echo=TRUE, warning=FALSE, message=FALSE}
doPlots(gpx, xVariable = "Time", doFacet = TRUE, sp = sp)
```

## Plot a comparison of the splits
```{r example12, echo=TRUE, warning=FALSE, message=FALSE}
plotCompareSplit(sp)
```

## Eliminate splits
```{r example13, echo=TRUE, warning=FALSE, message=FALSE}
gpx <- intervalActivity
sp <- autoSplits(gpx)
doPlots(gpx, xVariable = "Time", doFacet = TRUE, sp = sp)
spReduced <- purgeSplits(sp, splitsToEliminate = c(1, 3, 5, 7, 9, 11, 12))
doPlots(gpx, xVariable = "Time", doFacet = TRUE, sp = spReduced)
```

## Decoupling
```{r example14, echo=TRUE, warning=FALSE, message=FALSE}
gpx <- evenActivity
decouplingRes <- decoupling(gpx, variable = "GAP", from = 10, to = 60)
decouplingRes$compareHR
decouplingRes$compareVar
decouplingRes$compareSlopes
```

# Technical description

## Loading data

The function dataLoader() loads the user provided TCX file and returns a data frame with as many rows as data points and as many columns as the variables found in the TCX file:


```{r dataLoader0, echo=TRUE, warning=FALSE, message=FALSE}
#Do not execute
# gpx <- dataLoader('path_to_file.tcx')
head(gpx, 5)
```

The data loader will also perform the following operation:

* Convert time to a minutes only format, starting from 0 at the beginning of the activity
* Deal with mis-matching variables due to missing data points in one or more channels
* Interpolate missing data point (i.e dropped power or heart rate for a number of seconds/minutes)
* Compute pace and speed
* Deal with long chunks of inactivity, i.e. bike ride coffee break

### Deal with mis-matching variables
In some activity (especially swimming), columns read from the tcx file can have different lengths. The package attempts to recover the missing values in the shorter columns. The missing rows are replaced by NAs. After this take place, the package will attempt to fill in the missing numbers.

## Replace missing numbers
Missing numbers are replaced following the rules:

* Missing number with row index = 1. The NA is replaced by a 0
* Missing number with index = length(variable column). The Na is replaced with the closest non NA value moving upwards into the variable column
* All other cases. Missing numbers are replaced using linear interpolation between the closest non NAs

The three cases are summarized in the following example

```{r replaceNAs, echo=TRUE, warning=FALSE, message=FALSE}
myVariable <- c(NA,1,2,3,NA,NA,NA,7,NA)
idx <- which(is.na(myVariable))
idxNoNa <- which(!is.na(myVariable))
#replace first NA
if (1 %in% idx){
  myVariable[1] <- 0
  idx <- idx[-1]
}
#replace last NA
if (length(myVariable) %in% idx){
  myVariable[length(myVariable)] <- myVariable[idxNoNa[length(idxNoNa)]]
  idx <- idx[-length(idx)]
}
#replace other NAs
if(length(idx>0)){
  newValues <- seq(from = myVariable[idx[1]-1],
                   to = myVariable[idx[length(idx)]+1],
                   length = length(idx)+2)
  myVariable[idx] <- newValues[seq(2,length(newValues)-1,1)]
}
myVariable
```

## Deal with long pauses
Long pauses can artificially inflate the duration of an activity. Such long pauses can occour, for instance, when stopping for red light, or when stopping for a coffee break. Even if the watch goes in auto-pause, or you press STOP, the time elapsed before the activity resumes is still counted towards the total activity elapsed time. This is because the time is not recorded as starting from 0 and then incrementing, and pausing while in STOP mode, but instead every event is recorded based on the timestamp. That is, if the activity starts at 8:00am, then you ride non stop for 2h before pressing the STOP button, then you have 30 min break, then you resume the ride for other 2h, the total time will be calculated as follow:

* Start time: 8:00am
* Pause start 10:00am
* Pause end: 10:30am
* Ride end: 12:30pm
* Total ride time: 12:30pm - 8:00am = 2.5h

This will clearly affect your pace/speed, any other average (Watts, cadence, HR etc) and the activity TSS.

The package will eliminate the pause by replacing the gap with the most frequent time interval for the rest of the activity (typically 1s). This replacement is not performed in pool-based swims, as doing so will force all the intervals (lengths) to be computed with the same time, making the activity an artificially perfectly constant swim. Open water swim do not have this problem, as the lapping/recording method is not based on pool lengths, but on standard time/distance mode, as for riding or cycling activities.

## Generate splits

### Automatic splitting
The easiest way to generate splits is to use the function *autoSplits()*. The function is based on how Garmin reacts to the Lap event during an activity, either a manually trigger lap or an automatic lap. When a lap is triggered, Garmin duplicates the set of data at the point of lapping as can be observed in this activity data frame extract:

```{r autoSp, echo=FALSE, warning=FALSE, message=FALSE}
gpx <- intervalActivity
idx <- which(diff(intervalActivity$Time)==0)
gpx[seq(idx[1]-3,idx[1]+3,1),seq(1,4,1)]
```

Identifying these points allows to easily splits an activity without having to provide extra-parameters to the function *createSplits()*

### Create splits
The function *createSplits()* is used to manually split the activity. Splits can be created based on time or distance, and can be created of all equal lengths in time or distance, or completely manual. Consider the following examples:
```{r manualSp, echo=TRUE, warning=FALSE, message=FALSE}
gpx <- evenActivity
sp <- createSplits(gpx, 2500, "everyKm")
doPlots(data = gpx, xVariable = "Distance", sp = sp)
sp <- createSplits(gpx, c(1500,3450,5000), "thisKm")
doPlots(data = gpx, xVariable = "Distance", sp = sp)
sp <- createSplits(gpx, 15, "everyMin")
doPlots(data = gpx, xVariable = "Time", sp = sp)
sp <- createSplits(gpx, c(3,12,30), "thisMin")
doPlots(data = gpx, xVariable = "Time", sp = sp)
```

### Assisted splitting
This function can assist to identify the exact point to use to generate splits. Consider the following example:

```{r assistedSp0, echo=TRUE, warning=FALSE, message=FALSE}
gpx <- intervalActivity
proposedSplits <- c(15, 17)
addSeparator(gpx, proposedSplits, xVariable = "Time", yVariable = "Distance")
```

In this kind of plot Time/Distance, every change of pace can be visualized as a change in the slope of the red line. The aim is to have the blue lines to cross the points where the change of slope can be seen:

```{r assistedSp1, echo=FALSE, warning=FALSE, message=FALSE}
sp <- autoSplits(gpx)
proposedSplits <- tcxAnalyser:::getSplitsValues(gpx, sp, "Time")
proposedSplits <- round(proposedSplits, digits = 1)
```

```{r assistedSp2, echo=TRUE, warning=FALSE, message=FALSE}
proposedSplits
addSeparator(gpx, proposedSplits, xVariable = "Time", yVariable = "Distance")
sp <- createSplits(gpx, proposedSplits, "thisMin")
doPlots(data = gpx, xVariable = "Time", sp = sp)
```

Other combinations of x/y variables can be chosen, such as Time/Watts. However, the x variable must be either 'Time' or 'Distance', as the *createSplits()* function requires the split values to be given in either 'Time' or 'Distance'.

## TSS

Training Stress Score (TSS) is a registered term brought to the sport community by TrainingPeaks, and developed by Dr. Andy Coggan and Hunter Allen. In simple term, it is a measure of how hard your workout was, considering your level of fitness and the duration of the workout. It is calculated in different ways for different sports and/or for different type of data available. Before discussing TSS, we need to define Functional Threshold Power (FTP).

### FTP

First of all a disclaimer. There are several definition of FTP, and for the sake of simplicity, here I adhere to the definition available in TrainingPeaks. Also, I will call FTP also the pace threshold for running (normally otherwise referred to FTPa) and the lactate heart rate threshold (LHRT). I will also not discuss how to measure them. There are plenty of resource available on-line (I suggest to look for the tests recommended by Joe Friel).

If TSS is a measure of a specific workout intensity, FTP is a measure of the overall level of fitness at a specific time of your preparation. For instance, it can be the average power you can output in 1h hard ride.

## Computing TSS

The general equation for TSS is:

$$TSS = 100 * \frac{s*IF^2}{3600}$$

where *s* is the activity time in seconds and IF is the Intensity Factor, defined as:

$$IF = \frac{NP}{FTP}$$

Where NP is the normalised power, and FTP is the functional threshold power.

The concept of normalised power was developed in cycling, but it can be extended to other metrics such as pace and heart rate. The normalised power is calculated as a rolling average over the activity, with a time step of 30s. That is, the activity is divided into a number of 30s intervals, and the normalised power is computed for all of them. This package will calculate the TSS for every 30s of activity, and then sum all together to give the total activity (or split) TSS.

A simple calculation shows that an activity of 1h where the NP is exactly the FTP has TSS equal to 100.

You might have notice that for running and swimming the FTP is given in pace. Because the harder the pace, the lower its value, the FTP and the NPs need to be converted to speed before computing the IF, otherwise one would have lower TSS for faster run of same duration. Consider for instance three activities of duration 3600s and performed by an athlete with FTP of 4:00min/km. The three activities has NP of 3:30min/km, 4:00min/km and 4:30min/Km respectively. One would expect the TSS to decrease going from the first to the last activity, as the pace decreases while all other parameters remain constant. However, if we use the raw data we have:

$$TSS_{1} = 100*\frac{3600*(\frac{3.5}{4})^2}{3600} = 76.56$$
$$TSS_{2} = 100*\frac{3600*(\frac{4}{4})^2}{3600} = 100$$
$$TSS_{3} = 100*\frac{3600*(\frac{3.5}{4})^2}{3600} = 126.56$$
To convert both FTP and NP to speed (km/h) from pace (min/km), we need to do 60/FTP and 60/NP. This gives:

$$IF_{pace} = \frac{\frac{60}{NP_{pace}}}{\frac{60}{FTP_{pace}}} = \frac{FTP_{pace}}{NP_{pace}}$$

With this in mind, we can now verify that the three activities give the expected trend of TSS:

$$TSS_{1} = 100*\frac{3600*(\frac{4}{3.5})^2}{3600} = 130.61$$
$$TSS_{2} = 100*\frac{3600*(\frac{4}{4})^2}{3600} = 100$$
$$TSS_{3} = 100*\frac{3600*(\frac{4}{4.5})^2}{3600} = 79.01$$

One final consideration must be made on the calculation of NP for running. The NP was introduced in cycling to normalise the effort put on different rides based on the power output, so that short and hard efforts can be compared with long and steady efforts. In running, we are all familiar with a simple fact: going uphill will slow you down, or at least will require more effort to keep the pace, and going down hill will make you faster. This said, how can we compare hilly runs with flat runs? This comparison is made possible with the introduction of the grade adjusted pace (GAP), also known as normalised graded pace (NGP). This metric, first introduced by C.T.M. Davies (The Physiological Responses to Running Downhill. Journal of Applied Physiology 32, 187–194 (1974)), allows to take into consideration the effect of hill. In his study, Davies found that on average runners slow down by 3.5% every 1% of uphill incline, while speed up by 1.8% every 1% of downhill incline.

Considering Davies findings, when calculating NP for the running pace, the pace is always adjusted with the course elevation profile a that point of the activity. This is possible only if the analysed file contains elevation information. If it does not, then the non-normalised pace is used.

### On comparing TSS results from this analyser and TrainingPeaks

The TSS calculated here are in general good agreement with the TSS calculated by TrainingPeaks. However, for some activity there can be a significant difference. This can be due to:

* My error. It is possible that there is some mistake in my equations. If you spot one, please let me know and I will fix it
* Different way of dealing with missing data points. I replace missing data with linearly interpolated ones. I have no way to tell how TrainingPeaks deals with missing data
* Different smoothing algorithm. I always compute the NP on 30s blocks of raw data. I do not know if TrainingPeaks discard some of the blocks or if it applies some smoothing filters to the data before computing NP
* Different NP calculation, especially for running pace and GAP. I use the numbers reported by Davis in the cited paper, however I am aware that other papers have been published reporting either different numbers or some particular cases, such as extreme slopes. I am aware that Strava implements some of these latest findings, such as the ones reported by A.E. Minetti (Energy cost of walking and running at extreme uphill and downhill slopes. Journal of Applied Physiology 93, 1039–1046 (2002))

## Decoupling
Decoupling is a useful technique to asses aerobic fitness. In a long and steady effort, typically in zone 2, for a fit athlete there is very little decoupling before the heart rate and the metric of comparison, typically pace, GAP, speed or power. That is, during the effort not only there is very little variation between the first half of the effort and the second half of the effort, but also the two metrics (heart rate and pace, GAP, speed or power) plotted against time will give a very flat line. Consider for instance the following activity:

```{r exampleDecoupling0, echo=TRUE, warning=FALSE, message=FALSE}
gpx <- evenActivity
doPlots(gpx)
```

It is a relatively long and steady run. Even at this level, the hear rate and the GAP appear to be quite flat. Let's take a closer look using the *decoupling()* function, and analysing the activity between 10 and 60 minutes:

```{r exampledecoupling1, echo=TRUE, warning=FALSE, message=FALSE}
gpx <- evenActivity
decouplingRes <- decoupling(gpx, variable = "GAP", from = 10, to = 60)
```

Note that the black vertical lines divides the activity in two halves of equal lengths.

The plot appears a lot less flat now, especially for the GAP. The difference between the minimum and the maximum gap is around 0.15min/Km, equivalent to 9s/Km. Also, the first half was faster than the second half, while the HR remains almost constant. This indicates poor aerobic fitness, and the misalignment between heart rate and GAP is called *decoupling*. Using the results from the function *decoupling()*, we can also put a number to the observed differences.

To visualise the variation in heart rate between the first and the second half we can look at the *compareHR* data frame. Ideally, we want to see that there is virtually no difference between the average HR for the first and the second half, and the average overall heart rate. We would also like to see that there is very little difference between the slope for the first and the second half, and both should be close to 0:

```{r exampledecoupling2, echo=TRUE, warning=FALSE, message=FALSE}
decouplingRes$compareHR
```

In this case it looks good. There is a variation of just 4bpm between the first and the second half, both slopes are very close to 0 and quite similar.

To visualise the variation in GAP between the first and the second half we can look at the *compareVar* data frame. Ideally, we want to see a similar behaviour as with the heart rate comparison table:
```{r exampledecoupling3, echo=TRUE, warning=FALSE, message=FALSE}
decouplingRes$compareVar
```

As we can see, there is a difference of about 0.12min between the two halves average, corresponding to roughly 7s/Km. Also, the two slopes are similar to each other, and most importantly they are not very close to 0. All these numbers support the initial observation of poor aerobic fitness.

Finally, we can compare the slopes of the two metrics: heart rate and GAP. In this table, ideally we are aiming for two very similar slopes, and both very close to 0:
```{r exampledecoupling4, echo=TRUE, warning=FALSE, message=FALSE}
decouplingRes$compareSlopes
```

Once again, we can see that the slope for the heart rate is relatively close to 0, as it should be in a steady effort. However, the GAP slope is not, in fact it is about 10 times higher than the slope for heart rate. This suggests poor aerobic fitness, measured through decoupling analysis. For more information about decoupling, head to the Joe Friel blog (link in the final section of this guide).

In all the slope comparisons, the raw data were normalised between 0 and 100 to be able to compare slopes from data of otherwise different  magnitude.

# Useful links

* TrainingPeaks on TSS

https://www.trainingpeaks.com/blog/what-is-tss/

https://help.trainingpeaks.com/hc/en-us/articles/204071944-Training-Stress-Scores-Explained

* On the effect of running up-hill (GAP)

https://link.springer.com/article/10.1007/BF00423214

* Analysis and use of cycling training data

https://link.springer.com/article/10.2165%2F11317840-000000000-00000

* Validation of mathematical model for cycling power

http://journals.humankinetics.com/doi/pdf/10.1123/jab.14.3.276

* Andy Coggan on training with power meter (has info about TSS, IF, NP)

https://www.researchgate.net/publication/237251583_Training_and_racing_using_a_power_meter_an_introduction

* Decoupling

https://www.trainingpeaks.com/blog/aerobic-endurance-and-decoupling/
